Biogeography of crop progenitors and wild plant resources in West Asia, 21--8.3 ka

### 2025-01-04 analysis/paper.tex
Optimise model code

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Open Quaternary}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources during
the transition to agriculture in West Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of {[}X{]} plant species known to have been gathered
or cultivated by early agricultural societies in Southwest Asia,
including the progenitors of the first crops. We used machine learning
to train an ecological niche model (ENM) of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using hindcast data from palaeoclimate simulations.
Species ranges were on average {[}X\%{]} {[}larger\textbar smaller{]} in
the Early Holocene compared to today, indicating {[}\ldots{]}. The
modelled ranges predict the observed occurrence of these species on
archaeological sites with {[}low\textbar medium\textbar high{]}
accuracy. The regional ubiquity of species in the archaeological record
is {[}not{]} correlated with the predicted size of its range and the
diversity of archaeobotanical assemblages is {[}not{]} correlated with
the predicted diversity of its environs. This indicates that trends in
taxonomic composition of the archaeobotanical record is {[}not{]} likely
to be influenced by environmental change and species turnover, not just,
as is often assumed, human economic choices.
\end{abstract}

@@

\begin{itemize}
\tightlist
\item[$\square$]
  Finish analysis:

  \begin{itemize}
  \tightlist
  \item[$\square$]
    Add terrain, soil
  \item[$\square$]
    Find a resolution that's workable
  \item[$\square$]
    Do thresholding?
  \end{itemize}
\item[$\square$]
  First draft:

@@

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\square$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials

    \begin{itemize}
    \tightlist
    \item[$\boxtimes$]
      Occurrence data
    \item[$\boxtimes$]
      Predictor data
    \item[$\boxtimes$]
      Random Forest model
    \end{itemize}
  \item[$\square$]
    Results

@@

  \begin{itemize}
  \tightlist
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-climate-periods
  \end{itemize}
\item[$\square$]
  Appendix with all hindcast predictions
\item[$\square$]
  References \& copyediting
\item[$\square$]
  Final proofread
\end{itemize}

@@

@@

Taxonomic names were resolves to the canonical form specified in the
GBIF Backbone Taxonomy (\textbf{GBIFSecretariat2023?}). So for example
occurrences for ``Bolboschoenus'' included all species and subspecies,
including specimens described as \emph{Bolboschoenus sp.},
\emph{Bolboschoenus maritimus} and \emph{Scirpus maritimus}. Domestic
species meeting our inclusion criteria were substituted with their wild
progenitor(s), where known.

@@

Occurrence data was obtained from the Global Biodiversity Information
Facility (GBIF) using via its application programming interface and the
R package rgbif (Chamberlain et al., 2024; Chamberlain and Boettiger,
2017). Although ENMs have reasonable predictive power even with small
training samples (Hernandez et al., 2006; Stockwell and Peterson, 2002;
Wisz et al., 2008), we excluded X taxa with less than fifty recorded
occurrences in our study region. We also excluded one taxon (\emph{Avena
sterilis}) with over 47,000 occurrences, as this would have been
computationally prohibitive and we were uncertain what account for such
a disproportionately high number of records.

@@

For hindcasting, we used reconstructed bioclimatic data for key periods
(\textbf{?@tbl-climate-periods}) generated from downscaled paleoclimate
simulations from the HadCM3 general circulation model (Brown et al.,
2018). Terrain and soil predictors were held constant, since
reconstructions of these variables in the past are not available at
sufficient scale. It is not likely that either macroscale topography or
soil characteristics have altered significantly over the period of time
considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

>>>>

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Open Quaternary}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in West
Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of 56 plant species found regularly in association
with early agricultural archaeological sites in West Asia, including the
progenitors of the first crops. We used machine learning to train an
ecological niche model of each species based on its present-day
distribution in relation to climate and environmental variables.
Predictions of the potential ranges of these species at key stages of
the Pleistocene--Holocene transition could then be derived from these
models using hindcast data from palaeoclimate simulations. Species
ranges are predicted to have been on average {[}X\%{]}
{[}larger\textbar smaller{]} in the Early Holocene compared to present
conditions, indicating {[}\ldots{]}. The modelled ranges predict the
observed occurrence of these species on archaeological sites with
{[}low\textbar medium\textbar high{]} accuracy. The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

\begin{itemize}
\tightlist
\item[$\boxtimes$]
  Finish modelling
\item[$\square$]
  First draft:

@@

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\boxtimes$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials
  \item[$\square$]
    Results

@@

  \begin{itemize}
  \tightlist
  \item[$\square$]
    tbl-occ-count
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-climate-periods
  \end{itemize}
\item[$\square$]
  Clean up package code
\item[$\square$]
  Appendix with all hindcast predictions
\item[$\square$]
  References \& copyediting
\item[$\square$]
  Final proofread
\end{itemize}

@@

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-region-1.pdf}}

}

\caption{\label{fig-region}Study region}

\end{figure}%

@@

Taxonomic names were resolves to the canonical form specified in the
GBIF Backbone Taxonomy (\textbf{GBIFSecretariat2023?}). So for example
occurrences for \emph{Bolboschoenus maritimus} also include those
recorded under the older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where known.

@@

\begin{verbatim}
Rows: 62 Columns: 5
-- Column specification --------------------------------------------------------
Delimiter: "\t"
chr (2): taxon, taxon_detail
dbl (3): n_present, p_present, n_occ

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

\begin{verbatim}
Warning: HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
\end{verbatim}

\begin{table}

\caption{\label{tbl-occ-count}Recorded occurrence of flora considered in
this study in Neolithic and contemporary West Asia}

\centering{

\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrr}
\toprule
 & \multicolumn{2}{c}{Occurrences} \\ 
\cmidrule(lr){2-3}
Taxon & Neolithic & Present \\ 
\midrule\addlinespace[2.5pt]
\emph{Triticum turgidum dicoccum}

incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 (44\%) & 200 \\ 
\emph{Hordeum spontaneum}

incl. \emph{Hordeum vulgare} & 62 (43\%) & 6912 \\ 
\emph{Triticum monococcum aegilopoides}

incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 (30\%) & 3213 \\ 
\emph{Bolboschoenus maritimus}

incl. \emph{Scirpus maritimus} & 33 (23\%) & 364 \\ 
\emph{Vicia ervilia} & 33 (23\%) & 858 \\ 
\emph{Buglossoides tenuiflora} & 31 (21\%) & 159 \\ 
\emph{Arnebia decumbens} & 25 (17\%) & 240 \\ 
\emph{Buglossoides arvensis} & 25 (17\%) & 251 \\ 
\emph{Medicago radiata} & 21 (14\%) & 883 \\ 
\emph{Androsace maxima} & 20 (14\%) & 130 \\ 
\emph{Vicia orientalis}

incl. \emph{Lens culinaris} & 17 (12\%) & 441 \\ 
\emph{Medicago astroites}

incl. \emph{Trigonella astroites} & 16 (11\%) & 130 \\ 
\emph{Arnebia linearifolia} & 14 (10\%) & 158 \\ 
\emph{Linum bienne}

incl. \emph{Linum usitatissimum} & 11 (8\%) & 274 \\ 
\emph{Gypsophila vaccaria}

incl. \emph{Vaccaria pyramidata} & 11 (8\%) & 326 \\ 
\emph{Carex divisa} & 10 (7\%) & 330 \\ 
\emph{Ficus carica} & 10 (7\%) & 4265 \\ 
\emph{Lathyrus oleraceus}

incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 (6\%) & 2269 \\ 
\emph{Vicia faba} & 8 (6\%) & 2578 \\ 
\emph{Aizoanthemopsis hispanica}

incl. \emph{Aizoon hispanicum} & 7 (5\%) & 303 \\ 
\emph{Bolboschoenus glaucus} & 7 (5\%) & 27\textsuperscript{\textit{*}} \\ 
\emph{Pistacia atlantica} & 7 (5\%) & 928 \\ 
\emph{Polygonum arenarium arenarium}

incl. \emph{Polygonum venantianum} & 7 (5\%) & 5\textsuperscript{\textit{*}} \\ 
\emph{Prosopis farcta} & 7 (5\%) & 1205 \\ 
\emph{Rumex pulcher} & 7 (5\%) & 794 \\ 
\emph{Ammi majus} & 6 (4\%) & 461 \\ 
\emph{Cicer reticulatum}

incl. \emph{Cicer arietinum} & 6 (4\%) & 193 \\ 
\emph{Hordeum bulbosum} & 6 (4\%) & 2036 \\ 
\emph{Polygonum corrigioloides} & 6 (4\%) & 4\textsuperscript{\textit{*}} \\ 
\emph{Salsola kali} & 6 (4\%) & 369 \\ 
\emph{Taeniatherum caput-medusae} & 6 (4\%) & 235 \\ 
\emph{Capparis spinosa} & 5 (3\%) & 4031 \\ 
\emph{Chenopodium album} & 5 (3\%) & 357 \\ 
\emph{Lolium temulentum} & 5 (3\%) & 148 \\ 
\emph{Poa bulbosa} & 5 (3\%) & 1401 \\ 
\emph{Aegilops crassa} & 4 (3\%) & 170 \\ 
\emph{Atriplex prostrata} & 4 (3\%) & 112 \\ 
\emph{Brachypodium distachyon} & 4 (3\%) & 1581 \\ 
\emph{Fumaria densiflora} & 4 (3\%) & 194 \\ 
\emph{Gypsophila pilosa} & 4 (3\%) & 79 \\ 
\emph{Hordeum murinum} & 4 (3\%) & 1342 \\ 
\emph{Lathyrus aphaca} & 4 (3\%) & 2085 \\ 
\emph{Lathyrus sativus} & 4 (3\%) & 297 \\ 
\emph{Secale cereale} & 4 (3\%) & 1153 \\ 
\emph{Triticum durum} & 4 (3\%) & 219 \\ 
\emph{Vitis sylvestris} & 4 (3\%) & 3\textsuperscript{\textit{*}} \\ 
\emph{Adonis flammea} & 3 (2\%) & 79 \\ 
\emph{Avena sterilis} & 3 (2\%) & 50001\textsuperscript{\textit{*}} \\ 
\emph{Bassia arabica} & 3 (2\%) & 230 \\ 
\emph{Bromus sterilis} & 3 (2\%) & 385 \\ 
\emph{Cephalaria syriaca} & 3 (2\%) & 144 \\ 
\emph{Glaucium aleppicum} & 3 (2\%) & 22\textsuperscript{\textit{*}} \\ 
\emph{Helianthemum ledifolium} & 3 (2\%) & 668 \\ 
\emph{Lolium rigidum} & 3 (2\%) & 1464 \\ 
\emph{Phalaris paradoxa} & 3 (2\%) & 370 \\ 
\emph{Quercus ithaburensis} & 3 (2\%) & 1431 \\ 
\emph{Triticum aestivum compactum} & 3 (2\%) & 209 \\ 
\emph{Verbena officinalis} & 3 (2\%) & 1046 \\ 
\emph{Vicia narbonensis} & 3 (2\%) & 1529 \\ 
\emph{Triticum urartu} & — & 2033 \\ 
\emph{Aegilops speltoides} & — & 1429 \\ 
\emph{Aegilops tauschii} & — & 1395 \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{*}}Excluded from modelling due to sample size\\
\end{minipage}

}

\end{table}%

Georeferenced occurrence data was obtained from the Global Biodiversity
Information Facility (GBIF) using via its application programming
interface and the R package `rgbif' (Chamberlain et al., 2024;
Chamberlain and Boettiger, 2017). GBIF occurrences marked has having
imprecise or duplicate coordinates were excluded from the training
dataset, as were fossil records. Although niche models have reasonable
predictive power even with small training samples (Hernandez et al.,
2006; Stockwell and Peterson, 2002; Wisz et al., 2008), we excluded 57
taxa with less than 50 occurrences in West Asia, following
recommendations for niche models generally and Random Forest-based
models specifically (Stockwell and Peterson, 2002;
\textbf{LuanEtAl2020?}). We also excluded one taxon (\emph{Avena
sterilis}) with over 50,000 occurrences, as this would have been
computationally prohibitive and we were uncertain what account for such
a disproportionately high number of records.

\begin{verbatim}
Rows: 35208 Columns: 11
-- Column specification --------------------------------------------------------
Delimiter: "\t"
chr (5): taxon, taxon_detail, scientific_name, genus, species
dbl (6): n_present, p_present, n_occ, gbif_key, longitude, latitude

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

@@

For hindcasting, we used reconstructed bioclimatic data for 5 key
periods (\textbf{?@tbl-climate-periods}) generated from downscaled
paleoclimate simulations from the HadCM3 general circulation model
(Brown et al., 2018). Terrain and soil predictors were held constant,
since reconstructions of these variables in the past are not available
at sufficient scale. It is not likely that either macroscale topography
or soil characteristics have altered significantly over the period of
time considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

\begin{verbatim}
Warning: Removed 51 rows containing missing values or values outside the scale range
(`geom_text_repel()`).
\end{verbatim}

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-model-accuracy-vs-n_occ-1.pdf}}

}

\caption{\label{fig-model-accuracy-vs-n_occ}Number of taxon occurrences
and model accuracy on test dataset}

\end{figure}%
